Present bias in exploratory choice

### 2017-01-28 doc/cogsci2017/paper.tex
add abstract and Exp 2 participants and result stats

@@

The problem of under-exploration seems remarkably common. One of the most severe
expressions of under-exploration is learned helplessness, in which an organism
experiences the absence of contol over the environment, learns that the
environment is uncontrollable, and thus ceases to take actions that might allow
it to regain its sense of control. Learned helplessness has been proposed to
underly most notably depression \citep{Abramson1978, Abramson1989} but also
problems ranging from difficulties in school CITE to poverty CITE.

While the cognitive appraisal of experienced events affects the development of
learned helplessness \citep{Abramson1978}, patterns of exploration clearly play
a role as well \citep{Huys2009, Teodorescu2014a}. In the case of depression,
interventions aimed at increasing the exploration of activities this might
produce a sense of pleasure and mastery have been found to be as effective as
those with a more cognitive orientation \citep{Jacobson1996}.

@@

% In ``observe or bet'' tasks in which people choose to either
% obtain information or reward on each trial, people appear to collect much more
% information than is needed \citep{tversky1966information}. But, in settings
% where the underlying reward probabilities might change, information-seeking
% appears close to optimal \citep{Navarro2016}.

@@

@@

As in Experiment 1, participants rated the video as more enjoyable () than the
slider task (), STAT.

@@

The results of the regression are shown in Figure~FIGNUM. As in Experiment 1,
participants are less likely to choose a new spinner when the current spinner
has a high value, $p>.999$. However, in this experiment we find no effect of
condition---participants were no more likely to explore a new spinner when there
was a temporal delay imposed between their choices and the received outcomes.

@@

>>>>

  Balancing exploration and exploitation is difficult, and under-exploration
  appears to be a particularly pressing problem. We propose that one possible cause of
  under-exploration is present bias, in which immediate rewards (like those
  gained from exploitation) loom larger than future rewards (like those
  gained from exploration). This possible cause of under-exploration is not
  addressed by past lab studies, in which choices generally yield token rewards
  that are converted to money at the end of the experiment, removing the
  inter-temporal aspect of the decision-making process. In this paper, we
  develop an exploratory choice task with immediately-consumed rewards. In
  Experiment 1, we show that people understand the task and respond as expected
  to standard decision variables. In Experiment 2, we introduce a condition in
  which there is a temporal delay between choices and outcomes, and test whether
  this increases exploration by reducing present bias. While we find no effect
  of the delay, we discuss plans for future experiments to test the validity and
  robustness of this result.

@@

Balancing exploitation of known options with exploration of known options
presents a difficult cognitive and motivational challenge. The problem of how to
trade off between these two conflicting goals has created whole fields of study
\citep{Mehlhorn2015, sutton1998reinforcement}, and in many cases achieving an
optimal balance is computationally intractible \citep{Guez2013}. Given these
facts, one might expect people to often deviate from optimal exploratory choice
in both directions, perhaps exploring too much in some situations but too little
in others.

Notably, though, while over-exploration does occur, a bias towards under-exploration in particular seems to be
observed across a wide range of domains. One of the most severe
expressions of under-exploration is learned helplessness, in which an organism
experiences the absence of contol over the environment, learns that the
environment is uncontrollable, and thus ceases to take actions that might allow
it to discover that it can in fact exert control. Learned helplessness has been proposed to
underly most notably depression \citep{Abramson1978, Abramson1989} but also
problems ranging from difficulties in school CITE to poverty CITE.
While the cognitive appraisal of experienced events affects the development of
learned helplessness \citep{Abramson1978}, patterns of exploration clearly play
a role as well \citep{Huys2009, Teodorescu2014a}. In the case of depression,
interventions aimed at increasing the exploration of activities this might
produce a sense of pleasure and mastery have been found to be as effective as
those with a more cognitive orientation \citep{Jacobson1996}.

@@

@@

Forty participants completed the experiment, which was conducted over
Amazon Mechanical Turk using the psiTurk framework \citep{Gureckis2015a}.
Participants were paid \$5.00 for their participation, with a performance-based
bonus of up to \$3.00. All participants received the full \$3.00 bonus.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/exp2results.pdf}
\caption{Model-based estimates of participants' probability of choosing a new
  spinner for different values of the current spinner in Experiment 2. Thick
  lines and shaded regions indicate the mean and 95\% posterior interval for the
  population-level parameters, while the thin lines indicate the mean posterior
  parameters for each of the 40 individual participants. Participants in the
  delayed-outcome condition were no more
  likely to explore at a given current-spinner value than those in the
  immediate-outcome condition.}
\label{fig:exp2results}
\end{figure}

As in Experiment 1, participants rated the video as more enjoyable on average (5.65 out of 7) than the
slider task (3.13 out of 7), $t(39)=9.26$, $p<.001$.

@@

The results of the regression are shown in Figure~\ref{fig:exp2results}. As in Experiment 1,
participants were less likely to choose a new spinner when the current spinner
has a high value, $p>.999$. However, in this experiment we found no evidence of
an effect of condition, with a posterior probability of $p=.558$ that the effect of condition was
greater than 0. This means that participants were no more likely to explore a new spinner when there
was a temporal delay imposed between their choices and the received outcomes.
However, there may be a small interaction between current spinner value and
condition, such that participants in the delayed condition were less sensitive
to the value of the spinner when making their choices. This might indicate that
the delayed condition was confusing to some participants, as a few individuals
(as seen in Figure~\ref{fig:exp2results}) changed their behavior very little
across current-spinner values. The posterior probability that this interaction
was above 0 is $p=.934$.

@@

In this paper we presented a novel task using immediately-consumed rewards that
may be used to study present bias in the context of exploratory choice. In
Experiment 1 we showed that participants appeared to understand the task and
respond reasonably to maximize their time in the video-watching task. They are
more likely to explore when the current exploitable option (current spinner) has
low value, and change their degree of exploration based on the range of possible
outcomes from choosing a new spinner. However, in Experiment 2 we found that
participants did not explore any more when there was a temporal delay between
choices and outcomes, even though we expected this delay to reduce the present
bias towards the greater immediate rewards of exploitation.

One explanation for our results is that people are in fact less affected by
present bias in the realm of exploratory choice than they are in other
inter-temporal choices.
